GMSE: an R package for generalised management strategy evaluation

### 2017-08-07 notebook/ms/ms.Rmd
Update GMSE model structure section

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011]. This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [e.g., managers, users; @Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and stakeholder behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017]. Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of managers and stakeholders.

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model manager and stakeholder decision-making. Genetic algorithms allow managers and stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.

@@

GMSE builds off of the traditional MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1). (1) A natural resources sub-model considers a population of discrete resources with individual traits (e.g., location, age) on a spatially-explicit landscape. This sub-model can include processes of resource birth, movement, interaction with the landscape, and death. The discrete nature of resource demographics naturally gives rise to stochasticity, and therefore uncertainty. Although the resource sub-model can interact with other sub-models, it is unique in not relying on them because ecological dynamics can be simulated in the absence of observation and management. (2) An observation sub-model simulates the process of data collection peformed by a manager to estimate resource abundance. Four types of data collection are permitted, including exhaustive resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and exhaustive sampling of the whole landscape iteratively (during which resources might move). Imperfect sampling from all of these mechanisms of data collections generates observation uncertainty. (3) A manager sub-model simulates a manager's analysis of collected data to estimate resource abundance, then compares this estimated abundance with the manager's target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for stakeholder actions (e.g., culling, scaring, etc.) that minimise resource deviation from the target abundance, as informed by the predicted consequences of stakeholder actions and stakeholder action histories. After an adaptive policy is found, (4) a user sub-model is run that allows each stakeholder to perform actions that affect resources or landscape cells. Stakeholders respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each stakeholder has found an adaptive strategy, stakeholder actions affect resources and landscape cells, feeding back into the resource sub-model.

```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework"}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
plot(x = 0, y = 0, type = "n", xlim = c(0, 100), ylim = c(0, 100), 
     xaxt = "n", yaxt = "n", xlab = "", ylab = "");
# Manager model 
manger_model_box <- mbox(x0 = 1, x1 = 30, y0 = 70, y1 = 99);
polygon(x = manger_model_box$x, y = manger_model_box$y, lwd = 3);
text(x = 15, y = 90, cex = 1.0, labels = "Manager");
text(x = 15, y = 83, cex = 1.0, labels = "model");
# User model
user_model_box <- mbox(x0 = 70, x1 = 99, y0 = 70, y1 = 99);
polygon(x = user_model_box$x, y = user_model_box$y, lwd = 3);
text(x = 85, y = 90, cex = 1.0, labels = "User");
text(x = 85, y = 83, cex = 1.0, labels = "model");
# Observation model
observe_model_box <- mbox(x0 = 1, x1 = 30, y0 = 1, y1 = 30);
polygon(x = observe_model_box$x, y = observe_model_box$y, lwd = 3);
text(x = 15, y = 23, cex = 1.0, labels = "Observation");
text(x = 15, y = 15, cex = 1.0, labels = "model");
# Resource model
resource_model_box <- mbox(x0 = 70, x1 = 99, y0 = 1, y1 = 30);
polygon(x = resource_model_box$x, y = resource_model_box$y, lwd = 3);
text(x = 85, y = 23, cex = 1.0, labels = "Natural");
text(x = 85, y = 15, cex = 1.0, labels = "resources");
text(x = 85, y = 7,  cex = 1.0, labels = "model");
# Game-theoretic (genetic algorithm) model
game_model_box <- mbox(x0 = 36, x1 = 65, y0 = 36, y1 = 65);
polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
text(x=50, y=58, labels="Genetic", cex=1.0);
text(x=50, y=51, labels="algorithm", cex=1.0);
sc1 <- seq(from = 0, to = pi, by = 0.001);
sc2 <- seq(from = pi, to = 2*pi, by = 0.001);
points(x = 2*cos(sc1) + 41, y = 42 + 2*sin(sc1), lwd = 1.5, type = "l");
points(x = 2*cos(sc2) + 41, y = 40 + 2*sin(sc2), lwd = 1.5, type = "l");
arrows(x0 = 43, x1 = 43, y0 = 42, y1 = 41, lwd = 1.5, length = 0.06);
arrows(x0 = 39, x1 = 39, y0 = 40, y1 = 41, lwd = 1.5, length = 0.06);
text(x=43, y=42, labels="Multiple", cex=0.7, pos = 4);
text(x=43, y=39, labels="Generations", cex=0.7, pos = 4);
# Arrows indicating direction of GMSE model
arrows(x0=15, x1=15, y0=30, y1=70, lwd=2, length=0.15);
arrows(x0=30, x1=70, y0=85, y1=85, lwd=2, length=0.15);
arrows(x0=85, x1=85, y0=70, y1=30, lwd=2, length=0.15);
arrows(x0=70, x1=30, y0=15, y1=15, lwd=2, length=0.15);
arrows(x0=30, x1=36, y0=70, y1=65, lwd=2, length=0.075, code=3);
arrows(x0=65, x1=70, y0=65, y1=70, lwd=2, length=0.075, code=3);
text(x=50, y=90, labels="Policy", cex = 1.0);
text(x=50, y=20, labels="Monitoring", cex = 1.0);
text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
text(x=90, y=52, labels="Actions", srt=-90, cex = 1.0);
```

>>>>

To effectively manage natural resources, an adaptive approach allows managers to iteratively update their models of resource dynamics and respond flexibly to changing conditions [@Keith2011]. This approach is especially effective when considering all aspects of the social-ecological system being managed, including the dynamics of resources, monitoring, and the decision-making processes of stakeholders [e.g., managers and users; @Bunnefeld2011; @Bunnefeld2014]. Management strategy evaluation (MSE) is a modelling framework, first developed in fisheries, for simulating all of these aspects of resource management in a way that uniquely considers the uncertainties inherent to every stage of the management process [@Bunnefeld2011; @Punt2016]. Nevertheless, MSE remains limited in its ability to model human decision-making [@Fulton2011a; @Dichmont2017]; manager decisions are typically based on simple rules, and user behaviour remains fixed over time instead of dynamically responding to changing resource availability and management decisions [@Schluter2012; @Melbourne-Thomas2017]. Here we introduce a generalised management strategy evaluation (GMSE), which incorporates a game-theoretic perspective to model the goal-oriented, dynamic decision-making processes of stakeholders.

The GMSE R package is a flexible, highly mechanistic, agent-based modelling tool to simulate all key aspects of natural resource management. GMSE considers a range of parameters to simulate resource dynamics and management policy options, and uses genetic algorithms to dynamically model stakeholder (manager and user) decision-making. Genetic algorithms allow stakeholders to find adaptive solutions to any simulated conditions given unique agent goals, allowing GMSE to model scenarios of conservation conflict.

@@

GMSE builds off of the MSE framework, which includes four sub-models, each of which runs once in a single time step of the broader model (Figure 1). (1) A population of discrete resources with individual traits (e.g., location, age) is modelled on a spatially-explicit landscape and can simulate resource birth, movement, interaction with the landscape, and death; the discrete nature of resources causes demographic stochasticity, and therefore uncertainty. This sub-model is unique in not relying on other sub-models because ecological dynamics can be simulated in the absence of observation and management. (2) Observation is modelled in one of four ways: resource counting on a subset of landscape cells [e.g., @Nuno2013], marking and recapturing a fixed number of resources, and resource counting on the whole landscape either one linear transect or one rectangular block at a time (during which resources might move). Sampling error from all of these mechanisms of observation generates a range of uncertainties depending on monitoring effort. (3) Managers analyse data collected from observations to estimate resource abundance, then compare this estimate with their target abundance. Policy is developed by calling the genetic algorithm (see below), which works within a manager's constraints to find costs for user actions on the resource (e.g., culling, scaring, etc.) that minimise deviation from the target abundance, as informed by the predicted consequences of each action on resource abundance and user action histories. After a suitable policy is found, (4) users can perform actions that affect resources or landscape cells. Users respond to policy individually, each calling the genetic algorithm to find actions that maximise their own utilities (e.g., maximise resource use or landscape yield) within their imposed constraints. Once each user has found an adaptive strategy, user actions affect resources and landscape cells, feeding back into the resource sub-model.

```{r, echo = FALSE, fig.height = 4, fig.width = 4, fig.align = 'left', fig.cap = "Description of one time step of the generalised management strategy evaluation framework, which is comprised of four separate sub-models."}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
plot(x = 0, y = 0, type = "n", xlim = c(0, 100), ylim = c(0, 100), 
     xaxt = "n", yaxt = "n", xlab = "", ylab = "");
# Manager model 
manger_model_box <- mbox(x0 = 1, x1 = 30, y0 = 70, y1 = 99);
polygon(x = manger_model_box$x, y = manger_model_box$y, lwd = 3);
text(x = 15, y = 90, cex = 1.0, labels = "Manager");
text(x = 15, y = 83, cex = 1.0, labels = "model");
text(x = 5, y = 73, labels="(3)");
# User model
user_model_box <- mbox(x0 = 70, x1 = 99, y0 = 70, y1 = 99);
polygon(x = user_model_box$x, y = user_model_box$y, lwd = 3);
text(x = 85, y = 90, cex = 1.0, labels = "User");
text(x = 85, y = 83, cex = 1.0, labels = "model");
text(x = 74, y = 73, labels="(4)");
# Observation model
observe_model_box <- mbox(x0 = 1, x1 = 30, y0 = 1, y1 = 30);
polygon(x = observe_model_box$x, y = observe_model_box$y, lwd = 3);
text(x = 15, y = 23, cex = 1.0, labels = "Observation");
text(x = 15, y = 15, cex = 1.0, labels = "model");
text(x = 5, y = 4, labels="(2)");
# Resource model
resource_model_box <- mbox(x0 = 70, x1 = 99, y0 = 1, y1 = 30);
polygon(x = resource_model_box$x, y = resource_model_box$y, lwd = 3);
text(x = 85, y = 23, cex = 1.0, labels = "Natural");
text(x = 85, y = 15, cex = 1.0, labels = "resources");
text(x = 85, y = 7,  cex = 1.0, labels = "model");
text(x = 74, y = 4, labels="(1)");
# Game-theoretic (genetic algorithm) model
game_model_box <- mbox(x0 = 36, x1 = 65, y0 = 36, y1 = 65);
polygon(x = game_model_box$x, y = game_model_box$y, lwd = 3);
text(x=50, y=58, labels="Genetic", cex=1.0);
text(x=50, y=51, labels="algorithm", cex=1.0);
sc1 <- seq(from = 0, to = pi, by = 0.001);
sc2 <- seq(from = pi, to = 2*pi, by = 0.001);
points(x = 2*cos(sc1) + 41, y = 42 + 2*sin(sc1), lwd = 1.5, type = "l");
points(x = 2*cos(sc2) + 41, y = 40 + 2*sin(sc2), lwd = 1.5, type = "l");
arrows(x0 = 43, x1 = 43, y0 = 42, y1 = 41, lwd = 1.5, length = 0.06);
arrows(x0 = 39, x1 = 39, y0 = 40, y1 = 41, lwd = 1.5, length = 0.06);
text(x=43, y=42, labels="Multiple", cex=0.7, pos = 4);
text(x=43, y=39, labels="Generations", cex=0.7, pos = 4);
# Arrows indicating direction of GMSE model
arrows(x0=15, x1=15, y0=30, y1=70, lwd=2, length=0.15);
arrows(x0=30, x1=70, y0=85, y1=85, lwd=2, length=0.15);
arrows(x0=85, x1=85, y0=70, y1=30, lwd=2, length=0.15);
arrows(x0=70, x1=30, y0=15, y1=15, lwd=2, length=0.15);
arrows(x0=30, x1=36, y0=70, y1=65, lwd=2, length=0.075, code=3);
arrows(x0=65, x1=70, y0=65, y1=70, lwd=2, length=0.075, code=3);
text(x=50, y=90, labels="Policy", cex = 1.0);
text(x=50, y=20, labels="Monitoring", cex = 1.0);
text(x=10, y=48, labels="Indicators", srt=90, cex = 1.0);
text(x=90, y=52, labels="Actions", srt=-90, cex = 1.0);
```
